---
title: "Data Wrangling in R"
subtitle: "Environment Set Up"
author: "[Your Name]"   # (!!!) Update your name
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    toc: true
    toc_float: true
---


<!-- #............................................................................. -->
<!-- # Environment Set Up -->
<!-- #............................................................................. -->

<hr>

```{r, echo=FALSE}
date_report <- Sys.Date()
date_due_pre <- as.Date("2025-10-08")
date_due_lecture5 <- as.Date("2025-10-08")
date_due_lecture6 <- as.Date("2025-10-09")
date_due_lecture7 <- as.Date("2025-10-10")
date_due_lecture8 <- as.Date("2025-10-10")
date_due_lecture9 <- as.Date("2025-10-11")
date_due_lecture10 <- as.Date("2025-10-11")
date_due_lecture11 <- as.Date("2025-10-13")
date_due_lecture6to10 <- as.Date("2025-10-13")
```

`r if(date_report <= date_due_pre) {"\\begin{comment}"}`
# Environment Set Up
`r if(date_report <= date_due_pre) {"\\end{comment}"}`

`r if(date_report > date_due_pre) {"\\begin{comment}"}`
# Environment Set Up (**Response needed by `r format(date_due_pre,"%B %d, %Y")`)
`r if(date_report > date_due_pre) {"\\end{comment}"}`

<br>
**Instructions**

* Update [Your Name] on the top
* No other changes are needed
* Click "Knit" on the toolbar. Packages required for Data Wrangling Letures should be automatically installed.
* DW_R_setup.html should be saved in the same folder as DW_R_setup.Rmd file

<br>

```{r setup, include=FALSE}

package_wishlist.V <- c(
  "tidyverse",             
  "readxl",                 # import xls, xlsx
  "openxlsx",               # import and format xlsx
  "haven",                  # import sas, spss, strata
  "DBI", "odbc", "RSQLite", # packages for SQL database connection
  "httr",                   # package for HTTP API
  "jsonlite",               # json parser in R
  "googlesheets4",          # package that simplifies API to Google sheet
  "WDI",                    # package that simplifies API to World Bank
  "rentrez",                # package that simplifies API to Entrez databases (the r-cran version is suspending due to cran checks)
  "VIM",                    # Visualizing and Imputing Missing Data
  "mice"                    # Multivariate Imputation by Chained Equations. One of the most powerful tools for handling missing data 
  )

package_installed.DF <- as.data.frame(installed.packages()) # packages already installed

package_toinstall.V <- package_wishlist.V[!package_wishlist.V %in% package_installed.DF$Package]
```


**`r length(package_wishlist.V)` packages** are required for the Data Wrangling Lectures: `r paste(package_wishlist.V, collapse=", ")`.         

You have **`r length(package_wishlist.V)-length(package_toinstall.V)`/`r length(package_wishlist.V)` packages** already installed.             

**`r length(package_toinstall.V)` package(s)** will be installed: `r paste(package_toinstall.V, collapse=", ")`            
 

```{r, echo=F, include=TRUE, results='asis'}
if(length(package_toinstall.V) > 0)
{
  for(ipkg in 1:length(package_toinstall.V))
  {
    pkgs <- package_toinstall.V[ipkg]
    cat("\n\n## Installing ",pkgs, "\n\n", sep="")
    install.packages(pkgs = pkgs, repos = "http://cran.us.r-project.org")
  }
}
```




<!-- ######################################### -->
<!-- ## DO NOT MAKE CHANGES AFTER THIS LINE ## -->
<!-- ######################################### -->

    


```{css, echo=FALSE}

h1 {
  font-size: 20px;
  font-weight: bold;
}

h2 {
  font-size: 18px;
  font-weight: bold;
}

p {
  font-family: Arial;
  font-size: 15px;
}

```